---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(tidyverse)
```

### Load data

```{r load-data}
load("brfss2013.RData")
```

* * *

## Part 1: Data

The data is collected for non-institutionalized adults 18 or older living in 
the  United States of America by randomly contacting subjects by phone, 
household land-line or cell-phone.  cell-phone interviews are collected for 
adults indicating they live in a private residence or college housing.

Data are collected annually for all 50 states, the District of Columbia, 
Puerto Rico, Guam and the U.S. Virgin.  And are collected at limited points 
in time for American Samoa, Federated States of Micronesia and Palau.

As the data are randomly collected sample from the full population of the 
U.S., results should be generalizable to the full adult population. The 
collection of demographic information provides the ability to stratify the data
to identify possible confounding factors such as geographic location, 
ethnicity related factors, etc.  However, as some collection is done via 
cell phone there is no way to validate that the subjects interviewed lives
at a private residence or college housing which could lead to some possible
out of scope subjects getting included.

In addition to data collected about the subjects, data about the condition,
time and other environmental factors are collected.   There are also a large
number of calculated fields consolidating information from multiple questions
or inferring new information based on other responses.

* * *

## Part 2: Research questions

**Research question 1:**
Is there a relationship between a subjects weight and the likelihood of having
had a heart attack or having been diagnosed with cardiovascular disease?

This information can be used to help people determine their risk of hear attack
based on their own weight.

**Research question 2:**
Are Veterans more likely to suffer from depression than non-veterans?

Can be used by the veteran administration to determine how much to allocate
for mental health services.

It can also be used to help veterans who might be suffering from a depressive
disorder to understand they are not alone so they will be more likely to 
get treated.

**Research question 3:**
Is there a relationship between Sugar Drinks use and rates of pre-diabetes and 
diabetes?

For health care professionals, if a correlation is identified this can be used
guide patients and possibly reduce instances of diabetes or for people who are
already diagnosed, if there is a correlation changing their consumption could
aid in treating the condition.

* * *

## Part 3: Exploratory data analysis

### Question 1

**Is there a relationship between a subjects weight and the likelihood of having
had a heart attack or having been diagnosed with cardiovascular disease?**

Step 1, get the metrics needed to answer the question, these are a subjects 
weight, height and cardiovascular history.  As a subjects weight is relative
to their height I will be using  (BMI)[https://www.cdc.gov/healthyweight/assessing/bmi/index.html] (X_bmi), which 
is a ratio of height to weight instead of just weight.  

Only keep data where we know the subject is a resident of the state they were
in when the subject was interviewed and that they live in a private residence.

```{r}
hrt_data_1 <- brfss2013 %>%
  select(X_state, pvtresd1, stateres, sex,
         cvdinfr4, cvdcrhd4, htm4, wtkg3, X_bmi5) %>%
  filter(pvtresd1 == "Yes" & 
           stateres == "Yes")

summary(hrt_data_1)
```
Remove unusually (probably erroneous) tall and short subjects and cases where 
the height is NA.

```{r}
hrt_data_1 <- hrt_data_1 %>%
  filter(!is.na(htm4) & htm4 < 500 & htm4 > 10)
```

Remove cases where the weight is NA and convert to decimal.  It is stored as
an integer with the last 2 digits being values to the right of the decimal point.

```{r}
hrt_data_1 <- hrt_data_1 %>%
  filter(!is.na(wtkg3)) %>%
  mutate(wtkg3 = wtkg3 / 100)

summary(hrt_data_1)
```

I'd say this looks good but I'm distressed by the large number of people over
100kg, that's, at best, really sad.  We still have some NA's in BMI

Next let's drop the NA's from Cardiovascular disease indicators and make them
boolean values.  Also, set a single cardio value which indicates whether an
subject has either had a heart attack or been diagnosed with caritvascular
disease.

I've also decided to calculate the BMI as picking apart the encoded one seems
overly complicated and there are more NA's than I would expect.

```{r}
hrt_data_1 <- hrt_data_1 %>%
  filter(!is.na(cvdinfr4) & !is.na(cvdcrhd4)) %>%
  mutate(cvdinfr4=cvdinfr4=="Yes",
         cvdcrhd4=cvdcrhd4=="Yes",
         cardio=cvdinfr4 | cvdcrhd4,
         bmi=wtkg3/(htm4/100)^2,
         pvtresd1=NULL,
         stateres=NULL,
         X_bmi5=NULL)

summary(hrt_data_1)
```

Add a factor indicating the relative level of weight based on the BMI.  The
threshold values are based on the definitions on the above lined CDC BMI
site.  I combined, low and normal weight and the 2 high obese categories to 
keep things simple.

```{r}
bmi_factors = c("low_to_normal", "overweight", "obese",  
                "severely to very severly obeses")
hrt_data_1 <- hrt_data_1 %>%
  mutate(bmi_category=case_when(
    bmi < 25 ~ bmi_factors[1],
    bmi < 30 ~ bmi_factors[2],
    bmi < 35 ~ bmi_factors[3],
    TRUE ~ bmi_factors[4]),
    bmi_category=factor(bmi_category,levels=bmi_factors, ordered=TRUE))

```

```{r}
hrt_data_1 %>%
  group_by(sex, cardio, bmi_category) %>%
  summarize(mean_bmi=mean(bmi),
            count=length(bmi)
            
  )
```
Accounting for the subjects sex, what are the counts of people with and without
cardiovascular conditions by BMI category and what percent of each group has
cardiovascular conditions.

```{r}
as.data.frame(table(hrt_data_1[,c("cardio", "bmi_category", "sex")])) %>%
  tidyr::pivot_wider(names_from=cardio, names_prefix="cardio_", values_from=Freq) %>%
  mutate(pct_cardio=(cardio_TRUE / (cardio_FALSE + cardio_TRUE)) * 100)
```

There appears to be a positive correlation between BMI and cardiovascular 
conditions for both men and women.

**Research quesion 2:**

**Are Veterans more likely to suffer from depression than non-veterans?**

Get the required data.  Filter for only subjects who indicated a private 
residence and were residents of the state they were in during the interview.
Also remove any cases where sex, veteran status and whether they had been
diagnosed with depression were not answered.

```{r}
vet_data_1 <- brfss2013 %>%
  select(X_state, pvtresd1, stateres, sex,
         veteran3, addepev2) %>%
  filter(pvtresd1 == "Yes" & 
           stateres == "Yes",
         !is.na(sex),
         !is.na(veteran3),
         !is.na(addepev2))

summary(vet_data_1)
```

```{r}
vet_data_1 %>% 
  select(sex, veteran3, addepev2) %>% 
  table() %>%
  as.data.frame() %>%
  tidyr::pivot_wider(names_from=addepev2, 
                     names_prefix="Depressed_", 
                     values_from=Freq) %>%
  mutate(pct_depressed=(Depressed_Yes/(Depressed_Yes + Depressed_No))*100)
```

Veterans do show a slightly higher incident of being depressed than non-veterans
for both males and females.  27.1% of Female veterans have been diagnosed with a
depressive disorder vs. 22.1% for non-vets.  The Male population also shows
a higher rate of diagnosis but only slightly at 14.4% vs. 13.8 for Males.

**Research quesion 3:**

**Is there a relationship between Sugar Drinks use and rates of pre-diabetes and 
diabetes?**

Get the required data.  Filter for only subjects who indicated a private 
residence and were residents of the state they were in during the interview.
Also remove any cases where the subject reported diabetes diagnosis during
pregnancy and where there was no answer (NA) for sugar sweetened drinks.

Convert to total count of drinks per month, for drinks per week multiply by 4
and for drinks per day multiply by 30.

Add together pop and non-pop sugar sweetened drinks to 1 value.

```{r}
# Convert drinks count to monthly equivalent
sugarCount <- function(value) {
  v <- lapply(value, function(x) {
    if ( x > 300 ) return((x - 300) / 30)
    if ( x > 200 ) return((x - 200) / 7)
    if ( x > 100 ) return(x - 100)
    return(x)
  }) 
  
  return(unlist(v))
}


pop_data_1 <- brfss2013 %>%
  select(X_state, pvtresd1, stateres, sex,
         prediab1, diabage2, 
         ssbsugar, ssbfrut2) %>%
  filter(pvtresd1 == "Yes" & 
           stateres == "Yes" &
           !is.na(sex) &
           !is.na(prediab1) & prediab1 != "Yes, during pregnancy" &
           !is.na(ssbsugar),
         !is.na(ssbfrut2)) %>%
  mutate(prediab1=prediab1 == "Yes",
         ssbs=sugarCount(ssbsugar),
         ssbf=sugarCount(ssbfrut2),
         sugar_drinks_per_day=ssbs+ssbf,
  )

summary(pop_data_1)
```
```{r}
pop_data_1 %>%
  ggplot(aes(x=sugar_drinks_per_day)) +
  geom_boxplot()
```

After filtering for pop, only 1 diabetes subject remained.  So, we will look 
at pre-diabetes diagnosis only.

The majority of data lies withing a small band at less 1 drink / day or less
on average.   The data will be grouped into XX groups, no sugar drinks, 1 drink
every other day (0.5 drinks per day), 1 drink per day, greater than 1 drink 
per day.

```{r}
drinkRate <- function(dr, factors) {
  
  v <- lapply(dr, function(x) {
    if (x == 0) return(factors[1])
    if (x <= 0.5) return(factors[2])
    if (x <= 1) return(factors[3])
    if (x <= 3) return(factors[4])
    return(factors[5])
  })
  
  return(factor(unlist(v), levels=factors))
}

pop_data_2 <- pop_data_1 %>%
  mutate(drink_cat=drinkRate(sugar_drinks_per_day, 
                             c("none", "every other day", "1 a day", 
                               "more than 1 less than 3 a day", "3 or more a day")))

summary(pop_data_2)
```

```{r}
pop_data_2 %>%
  group_by(sex, prediab1, drink_cat) %>%
  summarize(count=n()) %>%
  tidyr::pivot_wider(names_from=prediab1, 
                     names_prefix="diagnosed_pre_diabetes_",
                     values_from=count) %>%
  mutate(percent_pre_diabetes=(diagnosed_pre_diabetes_TRUE /
                                 (diagnosed_pre_diabetes_TRUE + diagnosed_pre_diabetes_FALSE)) * 100)

```  

For males, there is a mostly negative correlation between number of sugary drinks
and diagnosis of pre-diabetes.  For women the ratio of population in each group
is relatively close.  Results are at best inconclusive.
